### 2024-07-02 latex/HP3_comm.tex
new input files

%\chapter*{Translation \& philological commentary}
%%%%%%%%%%
\subsection*{3.1}
\begin{translation}[hp03_001]
Just as the lord of snakes is the foundation of the regions of the earth along with their mountains and forests, so \emph{kuṇḍalinī} is the foundation of all systems of yoga.
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.2}
\begin{translation}[hp03_002]
When the sleeping \emph{kuṇḍalinī} awakens through the favour of the guru, then all the lotuses are pierced, and the knots too, [\dots] 
\end{translation}

@@

\begin{translation}[hp03_004]
\emph{Suṣumṇā}, the empty pathway, the great path to the aperture of Brahmā, the cremation ground, Śāmbhavī, and the middle path are synonyms.
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.6}
\begin{translation}[hp03_006]
The great seal, the great lock, the great piercing, the sky-roving [seal], the \emph{uḍḍiyāna} [lock], the root lock and the lock called \emph{jālandhara}, [\dots] 
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.12}
\begin{translation}[hp03_012]
[\dots] so the goddess \emph{kuṇḍalinī} suddenly becomes straight. Then she becomes still in the vessel with two halves.
\end{translation}

@@

\emph{Yogacintāmaṇi} ff. 72v–73r (attr.~to the \emph{Haṭhapradīpikā}) % order of the hemistichs has been changed. I don't know if or where I should mention this.
\begin{versinnote}
\tl{ṛjvībhūtā tathā śaktiḥ kuṇḍalī sahajā bhavet |\\+}
\tl{tathāsau maraṇāvasthāṃ harate dvipuṭāśrayām ||\\+}
\tl{dvipuṭā ] em., dvipaṭā° L, dvipadā° N. \\!}
\end{versinnote}

@@

\begin{philcomm}[hp03_012]
This verse is taken from the \emph{Amaraugha}, which uses the alchemical imagery of the \emph{Amṛtasiddhi} to describe the stilling of \emph{kuṇḍalinī} in the central channel. Drawing on Hellwig 2009:238–240, Mallinson and Szanto (2021:21) note that “In alchemical texts \emph{māraṇa} (“killing”) involves heating a substance and thereby changing its state, usually through calcination or oxidation, so that it becomes inert. In the \emph{Amṛtasiddhi māraṇa} and other derivatives of the root \emph{mṛ}, “die”, are used to denote the stilling or stopping of either the breath or Bindu.” Thus when \emph{kuṇḍalinī} is said to be in the state of \emph{maraṇa} the meaning is that she is stilled. The \emph{dvipuṭa} or “vessel with two halves” in which this occurs is the same as the \emph{Amṛtasiddhi}’s \emph{saṃpuṭa}, which, drawing on Hellwig 2009:342, Mallinson and Szanto (2021:22) say “consists of two \emph{puṭa}s joined together to form a sealed crucible for heating reagents without evaporation”. In the yoga of the \emph{Amṛtasiddhi}, the bodily \emph{saṃpuṭa} is formed by applying locks at the top and bottom of the central channel, i.e.~constricting the perineal region and the throat. In the \emph{Haṭhapradīpikā} it is formed by pressing the perineum with the heel and constricting the throat.

As Birch (2019: 971) notes, it is unlikely that later non-Buddhist authors understood \textit{maraṇāvasthā} and \emph{dvipuṭa} according to the alchemical metaphors of the \textit{Amṛtasiddhi}. Later commentators take \emph{dvipuṭa} as the two nostrils (e.g., \textit{Yuktabhavadeva} 7.187, \textit{dvināsāpuṭa}); the \textit{iḍā} and \textit{piṅgalā} channels (e.g., \textit{Jyotsnā} 3.27, \textit{puṭayor dvayam iḍāpiṅgalayor yugmam}); or the in and out flows of the breath (e.g., \textit{Yogaprakāśikā} 5.16–17, \textit{vāyor bahirnirgamanam antaḥpraveśa iti yat puṭadvayaṃ tam}). How these commentators understood \textit{asau māraṇāvasthā} is less clear. Brahmānanda seems to take it as the death of \textit{prāṇa}, or in other words, the absence of the breath, in the two nostrils (\textit{maraṇāvasthā jāyate kuṇḍalībodhe sati suṣumnāyāṃ praviṣṭe prāṇe dvayoḥ puṭayoḥ prāṇaviyogāt}).  Bhavadeva thought that \emph{kuṇḍalinī}, along with \textit{prāṇa} and \textit{apāna}, remains in the two nostrils while the breath is being held (\textit{evaṃ vāyudhāraṇāyāṃ kriyamāṇāyāṃ vyākulā bhūtā kuṇḍalinī apānaprāṇābhyāṃ saha nāsāpuṭadvayāśritā bhavati}). Others, such as Śivānanda and Bālakṛṣṇa, favour the reading \textit{tadā sā maraṇāvasthāṃ harate dvipuṭāśritām} (or \textit{dvipuṭāśrayām}), which is present in group 2 and 3 manuscripts of the \textit{Haṭhapradīpikā}. Bālakṛṣṇa understands this to mean that the great seal destroys death (\textit{maraṇāvasthāṃ harate mahāmudreti bhāvaḥ}) but it could also mean that the awakened \emph{kuṇḍalinī} destroys death, which is usually dependent on the in and out breaths.

@@

%%%%%%%%%%
\subsection*{3.13}
\begin{translation}[hp03_013]
{}[The yogi] should then exhale very slowly, not quickly. This is the great seal revealed by the great Siddhas.
\end{translation}

@@

\begin{philcomm}[hp03_025]
This verse, which is from the \emph{Amaraugha} and similar to verses in the \emph{Amṛtasiddhi} and \emph{Śivasaṃhitā}, is stating that the great seal, lock and piercing should be practised together. This can be done as a sequence as shown in this video. %?? insert video here
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{3.33}
\begin{translation}[hp03_033]
Now, the sky-roving [seal] (\emph{khecarī}):\\
By cutting, moving and milking, the yogi should gradually lengthen the tongue until it touches the middle of the brows. Then there is success in \emph{khecarī}.
\end{translation}

@@

\begin{metre}[hp03_033]
Upagīti 
\end{metre}

@@

%%%%%%%%%%
\subsection*{3.35}
\begin{translation}[hp03_035]
If the yogi turns back the tongue and remains [like that] for half an instant, he is instantly freed from disease, death, old age and the like.
\end{translation}

\begin{sources}[hp03_035]
\emph{Śivasaṃhitā} 3.91
\begin{versinnote}
\tl{rasanām ūrdhvagāṃ kṛtvā kṣaṇārdhaṃ yadi tiṣṭhati |\\+}
\tl{kṣaṇena mucyate yogī vyādhimṛtyujarādibhiḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp03_035]
\emph{Yogacintāmaṇi} f.~74r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{kalāṃ parāṅmukhīṃ kṛtvā tripathe parivartayet |\\+}
\tl{sā bhavet khecarī mudrā vyomacakraṃ tad ucyate ||\\+}
\tl{rasanām ūrdhvagāṃ kṛtvā kṣaṇārdhaṃ yadi tiṣṭhati |\\+}
\tl{viṣayair mucyate yogī vyādhimṛtyujarādibhiḥ ||\\!}
\end{versinnote}

@@

\begin{philcomm}[hp03_035]
We have adopted the version of 2.35 in \alphaTwo\ and \alphaThree\ (\alphaOne\ omits 2.35 and the greyscaled verses). This version is also attested by the \textgamma\ and \texteta\ groups. The additional hemistich gives \emph{vyomacakra} as an alternative name for \emph{khecarīmudrā}. This alternative name does not occur in any of the source texts known to be have been used by Svātmārāma, but it may have been inspired by the name \emph{nabhomudrā} for \emph{khecarīmudrā} in the \emph{Vivekamārtaṇḍa} (40).   
\end{philcomm}

@@

\begin{sources}[hp03_037]
\emph{Vivekamārtaṇḍa} 49
\begin{versinnote}
\tl{pīḍyate na sa rogeṇa lipyate na ca karmaṇā |\\+}
\tl{bādhyate na sa kālena yo mudrāṃ vetti khecarīṃ ||\\!}
%JB: Retain variants for lipyate na ca as alpha reads na ca lipyata/lipyati
\end{versinnote}
\end{sources}

@@

%\begin{philcomm}[hp03_037]
%\end{philcomm}

@@

%%%%%%%%%%
\subsection*{3.40}
\begin{translation}[hp03_040]
Even when semen has moved [down] and reached the region of the perineum, it moves upwards having been blocked by \emph{yonimudrā} and struck by the goddess [\emph{kuṇḍalinī}].
\end{translation}

@@

\tl{53c \var{gacchaty ] vrajaty \vl 53d hataḥ ] tanaṃ, hṛtas, kṛte, kṛtaḥ, tadā, gatā \vl • nibaddho ] niruddho \vl}\\!}
\end{versinnote}

@@

A manuscript of an early recension of the \emph{Vivekamārtaṇḍa} (ms.~no.~4110) has \emph{hataḥ śaktyā}, which is attested by three \emph{Haṭhapradīpikā} manuscripts on lower branches of the stemma (i.e.~\getsiglum{Ba} of \textdelta\ group and \getsiglum{C2} and \getsiglum{P4} of group 7a ??). The participle \emph{haṭaḥ} makes sense of the instrumental \emph{śaktyā}, rendering the meaning `struck by \emph{kuṇḍalinī}.' 
% MD: These groups don't have a Greek sigla yet.

Alternatively, the word \emph{haṭaḥ} (as well as the other variants \emph{kṛtaḥ, kṛte}, and even \emph{haṭhāt}) may derive from \emph{hṛtaḥ}, which is attested by \emph{Haṭhapradīpikā} manuscripts from group 4a (i.e., \getsiglum{Ba5}, \getsiglum{N4a} and \getsiglum{N13}). The reading \emph{hṛtaḥ śaktyā} renders the verse as saying that semen goes up, carried by \emph{kuṇḍalinī}.

@@

% write a note on yonimudrā
In \emph{Jyotsnā} 3.43, Brahmānanda explains `\emph{yonimudrā}' as `taking the form of contracting the penis' (\emph{yonimudrayā meḍhrākuñcanarūpayā}). He may have had in mind the practice of contracting and drawing the urethra upwards, which is described below in the section on \emph{vajrolimudrā} (\emph{Haṭhapradīpikā} 3.82). The author of the \emph{Yogaprakāśikā} (5.66) says that \emph{yonimudrā} is well-known in treatises on \emph{mantra} (\emph{yonimudrayeti mantraśāstraprasiddhayety arthaḥ} | \emph{°prasidhyayety} ed.). This is consistent with the \emph{Śivasaṃhitā}'s teaching on \emph{yonimudrā}, where it is described as activating the perineum (\emph{yoni}) by contracting it (4.2) to bring about success in mantra repetition (5.12). Later compendiums on yoga reiterate the role of \emph{yonimudrā} in mantra practice (e.g., the \emph{Yogacintāmaṇi} f.~65r, citing the \emph{Pārameśvaratantra}, and \emph{Haṭhatattvakaumudī} 33.12). The \emph{Haṭhayogasaṃhitā} (43–48) teaches a different version of \emph{yonimudrā} in its reportoire of twenty-five \emph{mudrā}s. In this work, \emph{yonimudrā} is supposed to awaken \emph{kuṇḍalinī} and consists of sitting in \emph{siddhasana}, blocking the ears, eyes, nose and mouth with the thumbs, index, middle and ring fingers, uniting \emph{praṇa} and \emph{apāna}, meditating on the six \emph{cakra}s and repeating the mantra \emph{huṃ haṃsa}.

@@

\begin{philcomm}[hp03_047]
The meaning of \emph{prāṇāt} (`from the breath') in the first verse quarter is not easy to understand without the context of this verse in the source text, the \emph{Vivekamārtaṇḍa}. In the verse preceding this one in the \emph{Vivekamārtaṇḍa} (117), the breath, on reaching the “great lotus”, is said to turn into nectar (\emph{amṛta}). In \emph{Jyotsnā} 3.51, Brahmānanda says that there is a variant \emph{prāṇaiḥ} (`by means of the breaths'), which is easier to understand than \emph{prāṇāt}. He nonetheless accepts \emph{prāṇat} and understands it as being a means (\emph{prāṇāt sādhanabhūtād avāptam}). He also understands the sixteen-petalled lotus to be the lotus in the throat, into which the nectar drips.
%Adopt patrapadma.
\end{philcomm}

\begin{metre}[hp03_047]
Śārdūlavikrīḍita 
\end{metre}

@@

%\begin{philcomm}[hp03_048]
%kaṭukāmla° or kaṭukātha. Adopt kaṭukātha (N3,J5, \emph{Vivekamārtaṇḍa})
%guṇavat (N3, J5 [-va])
%MD: Done.
%\end{philcomm}

@@

The first verse quarter has many variants among the witnesses. \alphaTwo\ has:
\begin{versinnote}
\tl{yat prāleyaṃ pihitasukhire merumūle yad astī \\+}
\tl{tasmiṃs tattvaṃ pravadati sudhīs tan mukhaṃ nimnagānām\\!}
\end{versinnote}
This version of the verse is close to that of other important manuscripts in the \textbeta, \textgamma\ and \texteta\ groups. With the help of these manuscripts (in particular \getsiglum{K1} of the \textgamma\ group), the above can be emended and understood as follows:
\begin{versinnote}
\tl{yat prāleyaṃ pihitasuṣiraṃ merumūrdhny asti tathyaṃ \\+}
\tl{tasmiṃs tattvaṃ pravadati sudhīs tan mukhaṃ nimnagānām\\+}
\tl{°suṣiraṃ ] \getsiglum{K1}, °sukhire \alphaTwo. °mūrdhny ] \getsiglum{P8}. °mūle \alphaTwo. asti tathyaṃ ] \getsiglum{K1}, yad astī \alphaTwo.\\!}
“That cool liquid by which the aperture is filled at the top of Meru and exists as the truth, the wise [yogi] says that is the source of [all] rivers.” 
\end{versinnote}

@@

%%%%%%%%%%
\subsection*{3.50}
\begin{translation}[hp03_050]
The root lock (\emph{mūlabandha})—\\
When [the yogi] presses the perineum with the side of the heel, clenches the anus and draws up \emph{apāna}, it is called the root lock.
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.51}
\begin{translation}[hp03_051]
[The yogi] forces the downward-moving \emph{apāna} breath to move upwards by contracting [the anus]. Yogis call that \emph{mūlabandha}.
\end{translation}

@@

In the context of the root lock, \emph{vāyum ākuñcayet} can be understood as an instruction to contract \emph{apāna\-vāyu}, which is mentioned in the previous verse (3.51a). Instructions to contract \emph{apāna\-vāyu} are found in other yoga texts, such as \emph{Yoga\-tārā\-valī} 7b (\emph{ākuñcanaiḥ śaśvad apānavāyoḥ}), \emph{Śiva\-saṃhitā} 4.84cd (\emph{apāna\-vāyum ākuñcya balād...}), \emph{Śiva\-yoga\-pradīpikā} 2.53ab (\emph{athordhva\-madhya\-sthira\-bandha\-nābhyām ākuñcanād ūrdhvam apānavāyoḥ}) and \emph{Yukta\-bhava\-deva} 7.297 (\emph{ādhāra\-kamale suptāṃ cālayet kuṇḍalīṃ dṛḍhām} | \emph{apāna\-vāyum ākṛṣya balād ākuñcya buddhimān}). In \emph{Jyotsnā} 3.63, Brahmānanda understands \emph{vāyu} in this verse as \emph{apāna} when he says that `one should contract the wind, \emph{apāna}' (\emph{vāyum apānam ākuñcayed}), which he explains as, `one should pull it by contractions of the anus' (\emph{gudasyā\-kuñcanenā\-karṣayet}). In the same vein, Bhava\-deva\-miśra glosses `pulling \emph{apāna}' (\emph{apānā\-karṣaṇam}) as `contracting the anus' (\emph{gudā\-kuñcanam}), when commenting on `having pulled \emph{apāna\-vāyu} and forcefully contracted it ...' (\emph{apāna\-vāyum ākṛṣya balād ākuñcya...}) in \emph{Yukta\-bhava\-deva} 297 and 301.  
\end{philcomm} %?? JB: I've added references to the commentators Brahmānanda and Bhavadevamiśra to clarify the meaning of vāyum ākuñcayet.

@@

%%%%%%%%%%
\subsection*{3.57}
\begin{translation}[hp03_057]
Heated by that [blaze], the sleeping \emph{kuṇḍalinī} wakes up. Like a snake struck by a stick, she hisses and becomes straight.
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.59}
\begin{translation}[hp03_059]
 Now the Uḍḍīyaṇa lock:\\
Because the breath flies up (\emph{uḍḍīyate}) into the \emph{suṣumṇā} when it is bound by it, yogis say that this [practice] is called Uḍḍīyaṇa.
\end{translation}

@@

\begin{philcomm}[hp03_061]
The variant spelling \sl{tāṇa} (for \emph{tāna}) reflects vernacular pronunciation (see e.g.~Molesworth 1857 s.v.~\emph{tāṇa}). 
\end{philcomm}

%%%%%%%%%%
\subsection*{3.62}
\begin{translation}[hp03_062]
\emph{Uḍḍiyāna} is easy, but it is always taught by a guru. If he practises it tirelessly, even an old man becomes young.
\end{translation}

>>>>

%\chapter*{Translation \& philological commentary}
%%%%%%%%%%
\subsection*{3.1}
\begin{translation}[hp03_001]
Just as the lord of snakes is the foundation of the regions of the earth along with their mountains and forests, so Kuṇḍalinī is the foundation of all systems of yoga.
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.2}
\begin{translation}[hp03_002]
When the sleeping Kuṇḍalinī awakens through the favour of the guru, then all the lotuses are pierced, and the knots too, [\dots] 
\end{translation}

@@

\begin{translation}[hp03_004]
Suṣumṇā, the empty pathway, the great path to the aperture of Brahmā, the cremation ground, Śāmbhavī, and the middle path are synonyms.
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.6}
\begin{translation}[hp03_006]
The great seal, the great lock, the great piercing, the sky-roving [seal], the \emph{uḍḍiyāna} [lock], the root lock, then [the lock] called \emph{jālandhara}, [\dots] 
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.12}
\begin{translation}[hp03_012]
[\dots] so the goddess Kuṇḍalinī suddenly becomes straight. Then she becomes still in the vessel with two halves.
\end{translation}

@@

\emph{Yogacintāmaṇi} ff. 72v–73r (attr.~to the \emph{Haṭhapradīpikā}) % order of the hemistichs has been changed. I don't know if or where I should mention this.
\begin{versinnote}
\tl{ṛjvībhūtā tathā śaktiḥ kuṇḍalī sahajā bhavet |\\+}
\tl{tathāsau maraṇāvasthāṃ harate dvipuṭāśrayām ||\\+}
\tl{\var{dvipuṭā ] em., dvipaṭā° L, dvipadā° N.} \\!}
\end{versinnote}

@@

\begin{philcomm}[hp03_012]
This verse is taken from the \emph{Amaraugha}, which uses the alchemical imagery of the \emph{Amṛtasiddhi} to describe the stilling of Kuṇḍalinī in the central channel. Drawing on Hellwig 2009:238–240, Mallinson and Szanto (2021:21) note that “In alchemical texts \emph{māraṇa} (“killing”) involves heating a substance and thereby changing its state, usually through calcination or oxidation, so that it becomes inert. In the \emph{Amṛtasiddhi māraṇa} and other derivatives of the root \emph{mṛ}, “die”, are used to denote the stilling or stopping of either the breath or Bindu.” Thus when Kuṇḍalinī is said to be in the state of \emph{maraṇa} the meaning is that she is stilled. The \emph{dvipuṭa} or “vessel with two halves” in which this occurs is the same as the \emph{Amṛtasiddhi}’s \emph{saṃpuṭa}, which, drawing on Hellwig 2009:342, Mallinson and Szanto (2021:22) say “consists of two \emph{puṭa}s joined together to form a sealed crucible for heating reagents without evaporation”. In the yoga of the \emph{Amṛtasiddhi}, the bodily \emph{saṃpuṭa} is formed by applying locks at the top and bottom of the central channel, i.e.~constricting the perineal region and the throat. In the \emph{Haṭhapradīpikā} it is formed by pressing the perineum with the heel and constricting the throat.

As Birch (2019: 971) notes, it is unlikely that later non-Buddhist authors understood \textit{maraṇāvasthā} and \emph{dvipuṭa} according to the alchemical metaphors of the \textit{Amṛtasiddhi}. Later commentators take \emph{dvipuṭa} as the two nostrils (e.g., \textit{Yuktabhavadeva} 7.187, \textit{dvināsāpuṭa}); the \textit{iḍā} and \textit{piṅgalā} channels (e.g., \textit{Jyotsnā} 3.27, \textit{puṭayor dvayam iḍāpiṅgalayor yugmam}); or the in and out flows of the breath (e.g., \textit{Yogaprakāśikā} 5.16–17, \textit{vāyor bahirnirgamanam antaḥpraveśa iti yat puṭadvayaṃ tam}). How these commentators understood \textit{asau māraṇāvasthā} is less clear. Brahmānanda seems to take it as the death of \textit{prāṇa}, or in other words, the absence of the breath, in the two nostrils (\textit{maraṇāvasthā jāyate kuṇḍalībodhe sati suṣumnāyāṃ praviṣṭe prāṇe dvayoḥ puṭayoḥ prāṇaviyogāt}).  Bhavadeva thought that Kuṇḍalinī, along with \textit{prāṇa} and \textit{apāna}, remains in the two nostrils while the breath is being held (\textit{evaṃ vāyudhāraṇāyāṃ kriyamāṇāyāṃ vyākulā bhūtā Kuṇḍalinī apānaprāṇābhyāṃ saha nāsāpuṭadvayāśritā bhavati}). Others, such as Śivānanda and Bālakṛṣṇa, favour the reading \textit{tadā sā maraṇāvasthāṃ harate dvi\-puṭāśritām} (or \textit{dvipuṭāśrayām}), which is present in group 2 and 3 manuscripts of the \textit{Haṭhapradīpikā}. Bālakṛṣṇa understands this to mean that the great seal destroys death (\textit{maraṇāvasthāṃ harate mahāmudreti bhāvaḥ}) but it could also mean that the awakened Kuṇḍalinī destroys death, which is usually dependent on the in and out breaths.

@@

%%%%%%%%%%
\subsection*{3.13}
\begin{translation}[hp03_013]
{[}The yogi] should then exhale very slowly, not quickly. This is the great seal revealed by the great Siddhas.
\end{translation}

@@

\begin{philcomm}[hp03_025]
This verse, which is from the \emph{Amaraugha} and similar to verses in the \emph{Amṛtasiddhi} and \emph{Śivasaṃhitā}, is stating that the great seal, lock and piercing should be practised together. This can be done as a sequence as shown in \medialink{this video}{the_three_mudrās.mp4} (© Mark Robberds).
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{3.33}
\begin{translation}[hp03_033]
Now, the sky-roving [seal] (\emph{khecarī}):\\
By cutting, moving, and milking, the yogi should gradually lengthen the tongue until it touches the middle of the brows. Then there is success in \emph{khecarī}.
\end{translation}

@@

\begin{metre}[hp03_033]
Āryā
\end{metre}

@@

%%%%%%%%%%
\subsection*{3.34*1}
\begin{translation}[hp03_034_1]
If the yogi turns back the tongue and remains [like that] for half an instant, he is instantly freed from disease, death, old age and the like.
\end{translation}

\begin{sources}[hp03_034_1]
\emph{Śivasaṃhitā} 3.91
\begin{versinnote}
\tl{rasanām ūrdhvagāṃ kṛtvā kṣaṇārdhaṃ yadi tiṣṭhati |\\+}
\tl{kṣaṇena mucyate yogī vyādhimṛtyujarādibhiḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp03_034_1]
\emph{Yogacintāmaṇi} f.~74r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{kalāṃ parāṅmukhīṃ kṛtvā tripathe parivartayet |\\+}
\tl{sā bhavet khecarī mudrā vyomacakraṃ tad ucyate ||\\+}
\tl{rasanām ūrdhvagāṃ kṛtvā kṣaṇārdhaṃ yadi tiṣṭhati |\\+}
\tl{viṣayair mucyate yogī vyādhimṛtyujarādibhiḥ ||\\!}
\end{versinnote}

@@

\begin{philcomm}[hp03_034_1]
This verse is absent in \alphaOne\ and was likely added to a block of verses that Svātmārāma borrowed from the \textit{Vivekamārtaṇḍa} (47–51). 3.34*1 is in \alphaTwo\ and \alphaThree\ (\alphaOne\ omits 2.35 and the greyscaled verses) and also the \textgamma\ and \texteta\ groups. Other manuscripts have an additional line that gives \emph{vyomacakra} as an alternative name for \emph{khecarīmudrā}. This alternative name does not occur in any of the source texts known to have been used by Svātmārāma, but it may have been inspired by the name \emph{nabhomudrā}, which is what the \emph{Vivekamārtaṇḍa} calls \emph{khecarīmudrā} (\emph{Vivekamārtaṇḍa} 40).   
\end{philcomm}

% (MD 2024-5-15): Should we perhaps take the absence of this verse in α1 seriously and regard it as inauthentic? HP 3.34-43 are all from VM, but not this 3.35. In α2 and α3 it may have been added independently by contamination. There are considerable differences in the readings between the two. JM: seems good to me.
% Pāda  α2 - α3
% e: nītvā - kṛtvā*
% f: api - yadi*
% g: viṣaye - kṣaṇena*
% The asterisk indicates that the reading has been adopted for our edition.

@@

\begin{sources}[hp03_037]
\emph{Vivekamārtaṇḍa} 49
\begin{versinnote}
\tl{pīḍyate na sa rogeṇa lipyate na ca karmaṇā |\\+}
\tl{bādhyate na sa kālena yo mudrāṃ vetti khecarīṃ ||\\!}
%JB: Retain variants for lipyate na ca as alpha reads na ca lipyata/lipyati
%MD(2024-5-14): Perhaps we should read na ca lipyati karmaṇā as epic Skt.
\end{versinnote}
\end{sources}

@@

\begin{philcomm}[hp03_037]
The form of \emph{lipyati}, which is found in the \alpha \ manuscripts and has been adopted, is widely attested in epic Sanskrit.
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{3.40}
\begin{translation}[hp03_040]
Even when semen has moved [down] and reached the region of the perineum, it moves upwards having been blocked by \emph{yonimudrā} and struck by the goddess [Kuṇḍalinī].
\end{translation}

@@

\tl{\var{53c gacchaty ] vrajaty \vl 53d hataḥ ] tanaṃ, hṛtas, kṛte, kṛtaḥ, tadā, gatā \vl • nibaddho ] niruddho \vl}\\!}
\end{versinnote}

@@

A manuscript of an early recension of the \emph{Vivekamārtaṇḍa} (ms.~no.~4110) has \emph{hataḥ śaktyā}, which is attested by three \emph{Haṭhapradīpikā} manuscripts on lower branches of the stemma (i.e.~\getsiglum{Ba} of \textdelta\ group and \getsiglum{C2} and \getsiglum{P4} of group 7a). The participle \emph{haṭaḥ} makes sense of the instrumental \emph{śaktyā}, rendering the meaning `struck by Kuṇḍalinī.' 
% MD: These groups don't have a Greek sigla yet.

Alternatively, the word \emph{haṭaḥ} (as well as the other variants \emph{kṛtaḥ, kṛte}, and even \emph{haṭhāt}) may derive from \emph{hṛtaḥ}, which is attested by \emph{Haṭhapradīpikā} manuscripts from group 4a (i.e., \getsiglum{Ba5}, \getsiglum{N4a} and \getsiglum{N13}). The reading \emph{hṛtaḥ śaktyā} renders the verse as saying that semen goes up, carried by Kuṇḍalinī.

@@

% write a note on yonimudrā
In \emph{Jyotsnā} 3.43, Brahmānanda explains `\emph{yonimudrā}' as `taking the form of contracting the penis' (\emph{yonimudrayā meḍhrākuñcanarūpayā}). He may have had in mind the practice of contracting and drawing the urethra upwards, which is described below in the section on \emph{vajrolimudrā} (\emph{Haṭhapradīpikā} 3.82). The author of the \emph{Yogaprakāśikā} (5.66) says that \emph{yonimudrā} is well-known in treatises on \emph{mantra} (\emph{yonimudrayeti mantraśāstraprasiddhayety arthaḥ} | \emph{°prasidhyayety} ed.). This is consistent with the \emph{Śivasaṃhitā}'s teaching on \emph{yonimudrā}, where it is described as activating the perineum (\emph{yoni}) by contracting it (4.2) to bring about success in mantra repetition (5.12). Later compendiums on yoga reiterate the role of \emph{yonimudrā} in mantra practice (e.g., the \emph{Yogacintāmaṇi} f.~65r, citing the \emph{Pārameśvaratantra}, and \emph{Haṭhatattvakaumudī} 33.12). The \emph{Haṭhayogasaṃhitā} (43–48) teaches a different version of \emph{yonimudrā} in its reportoire of twenty-five \emph{mudrā}s. In this work, \emph{yonimudrā} is supposed to awaken Kuṇḍalinī and consists of sitting in \emph{siddhasana}, blocking the ears, eyes, nose and mouth with the thumbs, index, middle and ring fingers, uniting \emph{praṇa} and \emph{apāna}, meditating on the six \emph{cakra}s and repeating the mantra \emph{huṃ haṃsa}.

@@

\begin{philcomm}[hp03_047]
The meaning of \emph{prāṇāt} (`from the breath') in the first verse quarter is not easy to understand without the context of this verse in the source text, the \emph{Vivekamārtaṇḍa}. In the verse preceding this one in the \emph{Vivekamārtaṇḍa} (117), the breath, on reaching the “great lotus”, is said to turn into nectar (\emph{amṛta}). In \emph{Jyotsnā} 3.51, Brahmānanda says that there is a variant \emph{prāṇaiḥ} (`by means of the breaths'), which is easier to understand than \emph{prāṇāt}. He nonetheless accepts \emph{prāṇat} and understands it as being a means (\emph{prāṇāt sādhanabhūtād avāptam}). He also understands the sixteen-petalled lotus to be the lotus in the throat, into which the nectar drips.
%Adopt patrapadma. % MD: VM-T has also padmapattra...
\end{philcomm}

% MD(2024-5-14): adopt vapur (α1,β1,ε2 etc.) instead of tanur (α2,β2,βω etc.)?
\begin{metre}[hp03_047]
Śārdūlavikrīḍita 
\end{metre}

@@

%\begin{philcomm}[hp03_048]
%MD(2024-5-14): adopt sadṛśā? (alpha and VM)
%\end{philcomm}

@@

The first verse quarter has many variants among the witnesses. \alphaTwo\ has:
\begin{versinnote}
\tl{yat prāleyaṃ pihitasukhire merumūle yad astī \\+}
\tl{tasmiṃs tattvaṃ pravadati sudhīs tan mukhaṃ nimnagānām\\!}
\end{versinnote}
This version of the verse is close to that of other important manuscripts in the \textbeta, \textgamma\ and \texteta\ groups. With the help of these manuscripts (in particular \getsiglum{K1} of the \textgamma\ group), the above can be emended and understood as follows:
\begin{versinnote}
\tl{yat prāleyaṃ pihitasuṣiraṃ merumūrdhny asti tathyaṃ \\+}
\tl{tasmiṃs tattvaṃ pravadati sudhīs tan mukhaṃ nimnagānām\\+}
\tl{\var{°suṣiraṃ ] \getsiglum{K1}, °sukhire \alphaTwo. °mūrdhny ] \getsiglum{P8}. °mūle \alphaTwo. asti tathyaṃ ] \getsiglum{K1}, yad astī \alphaTwo.}\\!}
“That cool liquid by which the aperture is filled at the top of Meru and exists as the truth, the wise [yogi] says that is the source of [all] rivers.” 
\end{versinnote}

@@

%%%%%%%%%%
\subsection*{3.50}
\begin{translation}[hp03_050]
The root lock (\emph{mūlabandha})—\\
When [the yogi] presses the perineum with part of the heel, clenches the anus and draws up \emph{apāna}, it is called the root lock.
\end{translation}
% JB: the back part of the heel presses the perineum. It would be better to say, 'with part of the heel.'

@@

%%%%%%%%%%
\subsection*{3.51}
\begin{translation}[hp03_051]
It forces the downward-moving \emph{apāna} breath to move upwards by contraction [of the anus]. Yogis call that \emph{mūlabandha}.
\end{translation}
% JB: Perhaps it is better to take MB as the subject of 51ab as MB is the subject of the previous verse? i.e., It forces the downward-moving \emph{apāna} breath to move upwards because of the contraction.

@@

In the context of the root lock, \emph{vāyum ākuñcayet} can be understood as an instruction to contract \emph{apāna\-vāyu}, which is mentioned in the previous verse (3.51a). Instructions to contract \emph{apāna\-vāyu} are found in other yoga texts, such as \emph{Yoga\-tārā\-valī} 7b (\emph{ākuñcanaiḥ śaśvad apānavāyoḥ}), \emph{Śiva\-saṃhitā} 4.84cd (\emph{apāna\-vāyum ākuñcya balād...}), \emph{Śiva\-yoga\-pradīpikā} 2.53ab (\emph{athordhva\-madhya\-sthira\-bandha\-nābhyām ākuñcanād ūrdhvam apānavāyoḥ}) and \emph{Yukta\-bhava\-deva} 7.297 (\emph{ādhāra\-kamale suptāṃ cālayet kuṇḍalīṃ dṛḍhām} | \emph{apāna\-vāyum ākṛṣya balād ākuñcya buddhimān}). In \emph{Jyotsnā} 3.63, Brahmānanda understands \emph{vāyu} in this verse as \emph{apāna} when he says that `one should contract the wind, \emph{apāna}' (\emph{vāyum apānam ākuñcayed}), which he explains as, `one should pull it by contractions of the anus' (\emph{gudasyā\-kuñcanenā\-karṣayet}). In the same vein, Bhava\-deva\-miśra glosses `pulling \emph{apāna}' (\emph{apānā\-karṣaṇam}) as `contracting the anus' (\emph{gudā\-kuñcanam}), when commenting on `having pulled \emph{apāna\-vāyu} and forcefully contracted it ...' (\emph{apāna\-vāyum ākṛṣya balād ākuñcya...}) in \emph{Yukta\-bhava\-deva} 297 and 301.  
\end{philcomm} %

@@

%%%%%%%%%%
\subsection*{3.57}
\begin{translation}[hp03_057]
Heated by that [blaze], the sleeping Kuṇḍalinī wakes up. Like a snake struck by a stick, she hisses and becomes straight.
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.59}
\begin{translation}[hp03_059]
Now the Uḍḍīyaṇa lock:\\
Because the breath flies up (\emph{uḍḍīyate}) into the Suṣumṇā when it is bound by it, yogis say that this [practice] is called Uḍḍīyaṇa.
\end{translation}

@@

\begin{philcomm}[hp03_061]
The spelling \sl{tāṇa} (where many witnesses have \emph{tāna}) reflects vernacular pronunciation (see e.g.~Molesworth 1857 s.v.~\emph{tāṇa}). 
\end{philcomm}

%%%%%%%%%%
\subsection*{3.62}
\begin{translation}[hp03_062]
\emph{Uḍḍiyāṇa} is easy, but it is always taught by a guru. If he practises it tirelessly, even an old man becomes young.%
\end{translation}
